---
title: '36662 Final Project: EDA Plots'
author: "Yuxi Chang"
date: "April 30, 2017"
output:
  pdf_document: 
    fig_caption: yes
    fig_height: 4
    fig_width: 5.5
  word_document: default
---

```{r, include=FALSE}
library(lattice)
library(knitr) # We need the knitr package to set chunk options

# Set default knitr options for knitting code into the report:
opts_chunk$set(echo = FALSE,# change to FALSE to keep code out of the knitted document
               tidy = TRUE,
               fig.align = "center",
               cache=TRUE, # do not re-run code that has already been run
               autodep=TRUE, # assure that caching dependencies are updated correctly
               cache.comments=FALSE, # do not re-run a chunk if only comments are changed
               message=FALSE, # change to FALSE to keep messages out of the knitted document
               warning=FALSE, # change to FALSE to keep warnings out of the knitted document
               comment = NA)
          
```

```{r data}
load("WeatherData.RData")

##Missing value
# there are 495 rows of data that response var DEP_DEL15 is missing. If using unsupervised data, we may
# create another dataset and give these rows value, so we will have more training/testing data.

# All missing PIT temp, humidity, visibility are all from the training data, create another version to 
# use the previous hour data for those NA's; 
# arrival data are missing 3 records, all from PHX, the new version uses the day before.

############ Features of Departure Airport (Pittsburgh) ################


```

#Flight Distance and Time

Flight Distance: Distances of most flights are in the range of 0-1000.\newline
Departure Time: 5-8am is the peak time. 15-17 is another peak time.\newline
Week Day: No big difference across week days.\newline
Yearly: No apparent seasonality across months.\newline


```{r distance and time}
### Flight Distance Distribution ###
par(mfrow = c(1,2))
histogram(tr15DEP$DISTANCE, xlab = "Flight Distance", main = "Flight Distance Distribution")

###Flight Time Distribution (By Hour)###

histogram(tr15DEP$DEP_Hour, xlab = "Departure Hour", main = "Departure Time Distribution")

### Weekly Flight Distribution ###

histogram(tr15DEP$DAY_OF_WEEK, xlab = "Departure Day", 
          main = "Departure Week Day Distribution")

### Monthly Trend ###
histogram(as.factor(tr15DEP$MONTH), xlab = "Departure Month", 
          main = "Departure Month Distribution")


```



#Features of Departure airport (Pittsburgh)

The heavier the snow, the lower the visibility.\newline

Rain is a little trickier; Level 2 has better visibility than Level 1?\newline

Heavier Rain/Snow, lower pressure.\newline


```{r section 1}


###Pittsburgh Humidity daily trend ###
par(mfrow = c(1,2))
dep_hum_hour <- aggregate(DEP_Humidity ~ DEP_Hour, data = tr15DEP, mean)
plot(DEP_Humidity ~ DEP_Hour, data = dep_hum_hour, type = "l", lwd = 1.5, 
     xlab = "Departure Hour", ylab = "Humidity", main = "Daily Humidity Trend")

### Pittsburgh Humidity Yearly trend ###
dep_hum_month <- aggregate(DEP_Humidity ~ MONTH, data = tr15DEP, mean)
plot(DEP_Humidity ~ MONTH, data = dep_hum_month, type = "l", lwd = 1.5, 
     xlab = "Month", ylab = "Humidity", main = "Yearly Humidity Trend")

###Pittsburgh Temperature daily trend ###
par(mfrow = c(1,2))
dep_temp_hour <- aggregate(DEP_Temp.F. ~ DEP_Hour, data = tr15DEP, mean)
plot(DEP_Temp.F. ~ DEP_Hour, data = dep_temp_hour, type = "l", lwd = 1.5, 
     xlab = "Departure Hour", ylab = "Temperature (F)", main = "Daily Temperature Trend")

### Pittsburgh Temperature Yearly trend ###
dep_temp_month <- aggregate(DEP_Temp.F. ~ MONTH, data = tr15DEP, mean)
plot(DEP_Temp.F. ~ MONTH, data = dep_temp_month, type = "l", lwd = 1.5, 
     xlab = "Month", ylab = "Temperature (F)", main = "Yearly Temperature Trend")

###Pittsburgh Pressure daily trend ###
par(mfrow = c(1,2))
dep_pres_hour <- aggregate(DEP_Pressure.in. ~ DEP_Hour, data = tr15DEP, mean)
plot(DEP_Pressure.in. ~ DEP_Hour, data = dep_pres_hour, type = "l", lwd = 1.5, 
     xlab = "Departure Hour", ylab = "Pressure (in.)", main = "Daily Pressure Trend")

### Pittsburgh Pressure Yearly trend ###
dep_pres_month <- aggregate(DEP_Pressure.in. ~ MONTH, data = tr15DEP, mean)
plot(DEP_Pressure.in. ~ MONTH, data = dep_pres_month, type = "l", lwd = 1.5, 
     xlab = "Month", ylab = "Pressure (in.)", main = "Yearly Pressure Trend")

###Pittsburgh Visibility daily trend ###
par(mfrow = c(1,2))
dep_vis_hour <- aggregate(DEP_Visibility.mi. ~ DEP_Hour, data = tr15DEP, mean)
plot(DEP_Visibility.mi. ~ DEP_Hour, data = dep_vis_hour, type = "l", lwd = 1.5, 
     xlab = "Departure Hour", ylab = "Visibility (mi.)", main = "Daily Visibility Trend")

### Pittsburgh Visibility Yearly trend ###
dep_vis_month <- aggregate(DEP_Visibility.mi. ~ MONTH, data = tr15DEP, mean)
plot(DEP_Visibility.mi. ~ MONTH, data = dep_vis_month, type = "l", lwd = 1.5, 
     xlab = "Month", ylab = "Visibility (mi.)", main = "Yearly Visibility Trend")

###Pittsburgh Wind-Speed daily trend ###
par(mfrow = c(1,2))
dep_wind_hour <- aggregate(DEP_Wind_Speed.mph. ~ DEP_Hour, data = tr15DEP, mean)
plot(DEP_Wind_Speed.mph. ~ DEP_Hour, data = dep_wind_hour, type = "l", lwd = 1.5, 
     xlab = "Departure Hour", ylab = "Wind Speed (mph)", main = "Daily Wind Speed Trend")

### Pittsburgh Wind-Speed Yearly trend ###
dep_wind_month <- aggregate(DEP_Wind_Speed.mph. ~ MONTH, data = tr15DEP, mean)
plot(DEP_Wind_Speed.mph. ~ MONTH, data = dep_wind_month, type = "l", lwd = 1.5, 
     xlab = "Month", ylab = "Wind Speed (mph)", main = "Yearly Wind Speed Trend")


### Pittsburgh Rain and Snow###
plot(DEP_Rain ~ as.factor(MONTH), data = tr15DEP, xlab = "Rain", main = "Yearly Rain", col = c("darkred", "darkblue","darkgreen","orange"))
plot(DEP_Snow ~ as.factor(MONTH), data = tr15DEP, xlab = "Snow", main = "Yearly Snow", col = c("darkred", "darkblue","darkgreen","orange"))

### Pittsburgh Snow/Rain vs. Visibility ###
snow_vis <- aggregate(DEP_Visibility.mi. ~ DEP_Snow, data = tr15DEP, mean)
plot(DEP_Visibility.mi. ~ DEP_Snow, data = snow_vis, ylab = "Visibility (mi.)", xlab = "Snow Level", main = "Visibility vs. Snow")
Rain_vis <- aggregate(DEP_Visibility.mi. ~ DEP_Rain, data = tr15DEP, mean)
plot(DEP_Visibility.mi. ~ DEP_Rain, data = Rain_vis, ylab = "Visibility (mi.)", xlab = "Rain Level", main = "Visibility vs. Rain")

### Pittsburgh Snow/Rain vs. Pressure ###
snow_pres <- aggregate(DEP_Pressure.in. ~ DEP_Snow, data = tr15DEP, mean)
plot(DEP_Pressure.in. ~ DEP_Snow, data = snow_pres, ylab = "Pressure (in.)", xlab = "Snow Level", main = "Pressure vs. Snow")
Rain_pres <- aggregate(DEP_Pressure.in. ~ DEP_Rain, data = tr15DEP, mean)
plot(DEP_Pressure.in. ~ DEP_Rain, data = Rain_pres, ylab = "Pressure (in.)", xlab = "Rain Level", main = "Pressure vs. Rain")


```

#Features of Destination Airports

There are several destination airports with data of only a few months.\newline

SFO, TPA, PBI, RSW, LAX, MCO, FLL have stable temperature.\newline

High Wind Speed: BOS, LGA, FLL, DAL\newline

Visibility: \newline
STL, MDW, MSP, ORD, PHL, IAH, DTW, EWR, HOU, ATL, BNA, BOS, BWI, CLT have ???low visibility during winter months. \newline

MCO, MDW, ATL, CLT, ??? High frequency of Thunder and Tornado 

```{r section 2}
### Destination Temperature ###
dest <- as.factor(tr15DEP$DEST)
month <- as.factor(tr15DEP$MONTH)
dep <- tr15DEP[,c(3,5,30:42)]
dep_grp <- aggregate(. ~DEST + MONTH, data = dep, mean)

dep_grp$ARV_Rain <- aggregate(as.numeric(tr15DEP$ARV_Rain), by = list(dest, month), FUN = sum)[,3]
dep_grp$ARV_Hail <- aggregate(as.numeric(tr15DEP$ARV_Hail), by = list(dest, month), FUN = sum)[,3]
dep_grp$ARV_Thunder <- aggregate(as.numeric(tr15DEP$ARV_Thunder), by = list(dest, month), FUN = sum)[,3]
dep_grp$ARV_Tornado <- aggregate(as.numeric(tr15DEP$ARV_Tornado), by = list(dest, month), FUN = sum)[,3]

xyplot(ARV_TEMP ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Temperature", main = "Destination Monthly Average Temperature")
xyplot(ARV_Pressure ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Pressure", main = "Destination Monthly Average Pressure")
xyplot(ARV_Visibility ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Visibility", main = "Destination Monthly Average Visibility")
xyplot(ARV_WindSpeed ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Wind Speed", main = "Destination Monthly Average Wind Speed")
xyplot(ARV_Precipitation ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Precipitation", main = "Destination Monthly Average Precipitation")
xyplot(ARV_Snowdepth ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Snow Depth", main = "Destination Monthly Average Snow Depth")
xyplot(ARV_Rain ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Rain", main = "Destination Monthly Rain")
xyplot(ARV_Hail ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Hail", main = "Destination Monthly Hail")
xyplot(ARV_Thunder ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Thunder", main = "Destination Monthly Thunder")
xyplot(ARV_Tornado ~ MONTH|DEST, data = dep_grp, pch = 19, ylab = "Tornado", main = "Destination Monthly Tornado")

```

#Carriers Distribution

```{r section 3}
histogram(tr15DEP$CARRIER, xlab = "Carrier", main = "Carrier Distribution")
table(tr15DEP$CARRIER, tr15DEP$DEST)
```

#Delay vs. Not Delay

No difference detected for distance and elapsed time. \newline
More delay flights during evening, less in the morning. \newline
Not delayed flights tend to have higher humidity.\newline
Not delayed flights tend to have lower wind speed.\newline
```{r section 4}
delay <- tr15DEP[which(tr15DEP$DEP_DEL15 ==1), ]
noDelay <- tr15DEP[which(tr15DEP$DEP_DEL15==0),]

# par(mfrow = c(1,2))
hist(delay$CRS_ELAPSED_TIME, main = "Elapsed time for delayed flights", xlab = "Elapsed Time", freq = FALSE)
hist(noDelay$CRS_ELAPSED_TIME, main = "Elapsed time for non-delayed flights", xlab = "Elapsed Time", freq = FALSE)

boxplot(delay$CRS_ELAPSED_TIME, delay$CRS_ELAPSED_TIME, main = "Delay vs. Not Delay: Elapsed time", names = c("delay", "noy delay"))

boxplot(delay$DISTANCE, noDelay$DISTANCE, main = "Delay vs. Not Delay: Distance", names = c("delay", "not delay"))

boxplot(delay$DEP_Hour, noDelay$DEP_Hour, main = "Delay vs. Not Delay: Departure Time", names = c("delay", "not delay"))

boxplot(delay$DEP_Temp.F., noDelay$DEP_Temp.F., main = "Delay vs. Not Delay: Temperature", names = c("delay", "not delay"))

boxplot(delay$DEP_Humidity, noDelay$DEP_Humidity, main = "Delay vs. Not Delay: Humidity", names = c("delay", "not delay"))

boxplot(delay$DEP_Pressure.in., noDelay$DEP_Pressure.in., main ="Delay vs. Not Delay: Pressure", names = c("delay", "not delay") )

delay_vis <- summary(delay$DEP_Visibility.mi.)
noDelay_vis <- summary(noDelay$DEP_Visibility.mi.)
vis <- cbind(delay_vis, noDelay_vis)
colnames(vis) <- c("Delay", "Not delay")
vis <- t(vis)
print(vis)

boxplot(delay$DEP_Wind_Speed.mph., noDelay$DEP_Wind_Speed.mph.,  main = "Delay vs. Not Delay: Wind Speed", names = c("delay", "not delay"))

boxplot(delay$ARV_Pressure, noDelay$ARV_Pressure, main = "Delay vs. Not Delay: Arrival Pressure", names = c("delay", "not delay"))

boxplot(delay$ARV_TEMP, noDelay$ARV_Pressure, main = "Delay vs. Not Delay: Arrival Temperature", names = c("delay", "not delay"))



```